---
title: "Does multi-omics integration earn its complexity?"
subtitle: "A worked introduction to `recoverR`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Does multi-omics integration earn its complexity?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4,
                      warning = FALSE, message = FALSE)
library(recoverR)
has_glmnet <- requireNamespace("glmnet", quietly = TRUE)
has_SNF    <- requireNamespace("SNFtool", quietly = TRUE)
set.seed(1)
```

## The two questions this package answers

Most multi-omics work reports that adding a layer *helped*, and the evidence is usually one of two
things: a variance share **inside a fitted model**, or a clustering at **one** parameterisation of
**one** integration algorithm. Neither tells you whether the extra layer would improve prediction on
data you have not seen, or whether the apparent gain would survive a different algorithm or a
different parameterisation.

`recoverR` supplies the missing tests:

1. **Attribution.** How much does each layer contribute to *out-of-sample* performance, and is that
   contribution distinguishable from a dimension-matched permutation null? `variancePartition`
   attributes variance to named covariates and `MOFA2` attributes it to latent factors; both are
   descriptive within a fitted model. `recoverR` decomposes a *predictive* functional instead.
2. **Recoverability.** Can a bulk profile carry a spatially-defined phenotype at all, and with what
   calibrated confidence? Reliability layers exist for *spatial* deconvolution output; this asks the
   inverse question, so a caller is told when the data cannot support the call they asked for.

Everything below runs on small synthetic data, so the vignette is fast. The two real-data calls are
shown but not evaluated; each is annotated with its measured cost.

## Part 1 — Attribution

### A small, complete example

We build four views that describe 120 samples with different amounts of shared signal. Views `v1` and
`v2` carry the outcome; `v3` is a noisy copy of `v1`; `v4` is pure noise.

```{r attribution-data}
make_view <- function(n, p, signal, noise) {
  base <- matrix(rnorm(n * p), n, p)
  if (signal) base[, 1] <- signal * y + rnorm(n)
  base + matrix(rnorm(n * p, sd = noise), n, p)
}
y <- rnorm(120)
X <- list(
  v1 = make_view(120, 30, signal = 1.0, noise = 1),
  v2 = make_view(120, 30, signal = 0.6, noise = 1),
  v3 = make_view(120, 30, signal = 0.0, noise = 1),
  v4 = make_view(120, 30, signal = 0.0, noise = 1)
)
sapply(X, dim)
```

`attrib_views()` returns the exact Shapley decomposition of out-of-sample performance, a
dimension-matched permutation null for each view, and leave-one-out / forward ablations.

```{r attribution-run, eval = has_glmnet}
res <- attrib_views(X, y, family = "gaussian", nfolds = 5L, seed = 1L,
                    n_perm = 49L, ncores = 2L, verbose = FALSE)
res
```

Read the two columns together:

- **`shapley`** is the view's total credit. It is *efficiency*-satisfying: the values sum to
  `V(full) - V(empty)`.
- **`unique_contribution`** is `V(full) - V(full \ k)`: what is lost when that view alone is removed.

A view whose Shapley is large while its unique contribution is near zero is earning credit for
performance **shared** with correlated views. In our example `v3` is deliberately a noisy copy of
`v1`, so it inherits a real share of `v1`'s credit — which is exactly why a bare variance share
overstates a redundant layer.

```{r attribution-null, eval = has_glmnet}
as.data.frame(res$null)[, c("view", "delta_observed", "null_mean", "null_q95", "p_value")]
```

The null is **dimension-matched by construction**: the permuted view has exactly the same number of
features as the real one, so no correction for dimensionality is needed. Its hypothesis is precise:

> H0: given the other views, this view carries no sample-specific information about `y`.

Note that a pure-noise view can carry a **non-zero** Shapley value — random columns act like
noise-injection regularisation. This is why a permutation null is needed rather than a bare R², and
why "assert p > 0.05 for a noise view" is a bad test: p is uniform under H0, so it fails about 5% of
the time by design. The stable discriminator is the **gap to the null mean**.

```{r attribution-plate, fig.height = 6, eval = has_glmnet}
plot_attribution_plate(res)
```

### What it costs on real data

A real multi-omics call looks like this. The cost depends on the
learner, and the difference is large enough to change how you design an analysis:

```{r real-calls, eval = FALSE}
# Four views, 300 features each, binomial endpoint, B = 49 permutations: ~3.5 min on 8 cores.
res_bin <- attrib_views(X_views, y_response, family = "binomial",
                        nfolds = 5L, n_perm = 49L, ncores = 8L)

# The SAME call with a Cox endpoint costs ~25 s per cross-validated fit against
# ~0.8 s for binomial, and the Shapley lattice is SERIAL, so a full-budget
# B = 199 run is ~2 hours - not ~10 minutes. Budget the estimator to the endpoint.
res_surv <- attrib_views(X_views, Surv(time, event), family = "survival",
                         nfolds = 3L, n_perm = 0L, ncores = 8L)
```

If a view survives the null, `ablate_views()` gives the leave-one-out and forward-selection views of
the same question, and `plot_leave_one_out()` / `plot_forward_selection()` draw them.

## Part 2 — Recoverability: can a bulk profile carry a spatial phenotype?

A bulk profile is an average. If a tumour is part inflamed and part desert, its average looks
moderately inflamed — a misleading summary. `recoverR` asks whether the archetype is recoverable at
all, and **abstains** when it is not.

The predictor sees only the aggregated expression of a region; the label comes from the spatial
ground truth. Nothing about the label is derived from the feature vector, so the test is not
circular: it measures precisely what aggregation destroys.

```{r conformal-data}
n <- 300
lab <- sample(c("inflamed", "excluded", "desert"), n, replace = TRUE)
shift <- c(inflamed = 1.2, excluded = -0.2, desert = -1.4)[lab]
Xr <- matrix(rnorm(n * 40), n, 40)
Xr[, 1] <- shift + rnorm(n, sd = 0.9)
Xr[, 2] <- shift + rnorm(n, sd = 1.1)
# Features must be NAMED: predict_recoverability() matches the calibration and
# prediction matrices by feature identifier, and an unnamed matrix shares no
# features with the training set (it reports "fewer than 2 shared features").
colnames(Xr) <- sprintf("g%02d", seq_len(ncol(Xr)))
table(lab)
```

```{r conformal-run}
fit <- refit_recoverability(Xr, factor(lab), alpha = 0.10, mondrian = TRUE,
                            score = "aps", cal_prop = 0.5, seed = 1)
c(calibration_coverage = fit$cal_coverage)
fit$cal_coverage_by_class
```

Split conformal prediction carries a finite-sample coverage guarantee: the prediction sets contain
the true archetype with probability at least `1 - alpha`, and **Mondrian** (class-conditional)
conformal makes that hold *per class* rather than only on average.

```{r conformal-predict}
pred <- predict_recoverability(fit, Xr)
as.data.frame(pred)$predicted[1:6]
mean(pred$abstain)
```

Abstention is the point: the method is allowed to say *"this profile cannot support a call"* rather
than emitting a confident archetype where the data do not warrant one.

```{r separability}
archetype_separability(fit)
```

The separability matrix names which archetype **pairs** the data cannot distinguish. In practice
that matrix, not the overall accuracy, is the actionable output: it says which clinical distinction a
bulk assay will not deliver.

```{r reliability, fig.height = 5}
plot_reliability(fit)
```

```{r coverage-risk, fig.height = 4}
plot_coverage_risk(fit)
```

`reliability_curve()` returns an expected calibration error, and `selective_gain()` reports what
abstention buys you in accuracy at a given coverage — the selective-risk view.

## Part 3 — Sampling adequacy: how much tissue, and how many genes?

Spatial signal collapses below a certain specimen size, and that constraint is usually reported once,
as an afterthought, for one study. `sampling_adequacy()` turns it into a measured quantity: at what
region diameter does the archetype stop being recoverable from the aggregate profile?

```{r adequacy, fig.height = 4.5}
set.seed(2)
n_spot <- 900
spots  <- sprintf("spot-%04d", seq_len(n_spot))
coords <- data.frame(x = rep(seq_len(30), each = 30) * 100,
                     y = rep(seq_len(30), times = 30) * 100,
                     barcode = spots, row.names = spots)

# A SMOOTH large-scale spatial field, so that a bigger region captures it more
# reliably than a small one - which is the saturating relationship the Hill fit
# expects. (A fine-grained pattern would instead average OUT with region size and
# the fit would be correctly rejected; see the notes on extrapolations below.)
field  <- sin(coords$x / 1200) + cos(coords$y / 1500)
labels <- setNames(ifelse(field > 0, "inflamed", "desert"), spots)

# counts must be SPOTS x GENES, with rownames matched to coords (see ?pseudobulk_regions)
cnt <- matrix(rpois(n_spot * 200, 5), nrow = n_spot,
              dimnames = list(spots, paste0("GENE", seq_len(200))))
cnt[,  1:25] <- cnt[,  1:25] + pmax(0,  field) * 6
cnt[, 26:50] <- cnt[, 26:50] + pmax(0, -field) * 6

sa <- sampling_adequacy(cnt, coords, labels = labels,
                        diameters = c(300, 600, 1200, 2400, 3600),
                        panel_sizes = c(50, 200), n_regions = 40L,
                        target = 0.95, seed = 1L)
sa$adequacy[, c("panel_genes", "plateau", "half_saturation_mm", "adequacy_mm", "extrapolated")]
```

Two things to read carefully:

- **`extrapolated`**. A Hill curve fitted to diameters that never reach its plateau cannot identify
  its half-saturation, and `d*` then falls *outside* the sampled range — an extrapolation, not a
  measurement. `recoverR` flags it so you report *"the required diameter exceeds what we measured"*
  rather than quoting a number.
- **The diameter is data-derived, not assumed.** Pass `diameters` in the units of your `coords`; the
  millimetre column in `$adequacy` is inherited from those units, so verify them. (A hard-coded
  pixel-to-micrometre constant is a classic silent error: the Visium spot *diameter* is 55 µm but the
  spot *pitch* is 100 µm, and the pixels-per-µm scale is image-dependent.)

```{r sampling-curve, fig.width = 7}
plot_sampling_curve(sa$surface)
```

### Reading real Visium data

```{r visium, eval = FALSE}
# One sample per section. The counts .h5 and the *_spatial.tar.gz are separate
# downloads and BOTH are required - the archive holds the coordinates.
v <- read_visium("path/to/visium/sample_dir", sample_id = "GSM0000000")

scores <- recoverR:::score_programmes(v$counts, rr_spatial_programmes())
sec    <- derive_spatial_archetypes(scores)          # section-level call
sec$archetype

# regions: pseudo-bulk expression plus the spatial ground truth for each region
pb <- pseudobulk_regions(t(v$counts), v$coords, diameter = 1500,
                         n_regions = 300, min_spots = 12, aggregate = "sum", seed = 1)
```

## Guardrails

`recoverR` wraps the failure modes that are silent rather than loud — the kind that return a
plausible number instead of an error.

```{r guardrails}
# NA in a feature matrix is usually a merge artefact, not biology
tryCatch(assert_no_na(matrix(c(1, NA, 3))), error = function(e) conditionMessage(e))

# a verbal check is not a check: confirm a PDF is genuinely vector, not rasterised
tmp <- tempfile(fileext = ".pdf")
grDevices::pdf(tmp); plot(1:10, 1:10); grDevices::dev.off()
is_vector_pdf(tmp)
```

| Helper | The silent failure it catches |
|---|---|
| `assert_no_na()` | NA introduced by a join, which propagates into every fit |
| `cv_glmnet_safe()` | `cv.glmnet` silently downgrading `type.measure = "auc"` to `"deviance"` when folds hold <10 observations, so `which.max(cvm)` then maximises *error* |
| `auc_safe()` | `pROC::roc.test` erroring at AUC = 1 and returning `NA`, which survives BH adjustment unnoticed |
| `is_vector_pdf()` | ggplot2 rasterising a continuous colourbar legend into a 1-px strip inside a "vector" PDF |
| `assert_palette_covers()` | a palette that silently drops a factor level, recolouring the wrong group |
| `safe_cbind()` | duplicate column names from per-layer PCA scores, which make `lm()` return `NA` |

## Publication figures

Every plotting function composes with `ggplot2`, and `save_pub()` writes **both** a vector PDF and a
JPEG at a journal-specified width, verifying that the PDF really is vector:

```{r figures, eval = has_glmnet}
plt <- plot_shapley(res)
out <- file.path(tempdir(), "shapley")
save_pub(plt, out, width_mm = 183, height_mm = 90)
basename(list.files(tempdir(), pattern = "^shapley", full.names = TRUE))
```

## Where to go next

- Every exported function carries a documented mathematical and behavioural description
  (`?attrib_views`, `?refit_recoverability`, `?sampling_adequacy`): the value functional per endpoint,
  the Shapley estimator and its efficiency argument, the permutation null, the conformal scores, and
  the Hill fit with its bound checks.
- The package's test suite (37 test blocks, 81 expectations) doubles as a set of worked
  micro-examples, including a synthetic Visium fixture written in the exact 10x layout.
- **Validation status, stated plainly.** Module 1 (attribution) and the estimator underneath it are
  regression-tested and were checked against planted signals with a known answer. Module 2's harness is
  verified by a positive control (it recovers a fully planted, zero-noise signal). What is **not** yet
  available to anyone, including us, is *independent ground truth* for the recoverability module: that
  requires spatially defined labels measured independently of the transcriptome (for example multiplex
  immunofluorescence on clinical material). Until then, read Module 2's real-data output as calibrated
  **relative to its own labels**, and read its abstention rate as a property of the data and the layer
  rather than a defect.
